---
title: "Impact of US Weather Events in 1996-2010"
author: "davegoblue"
date: "February 16, 2016"
output: html_document
---

## Synopsis  
This analysis investigates the economic and health impact of US weather events based on an NOAA database spanning 1950-2011.  Weather event tracking in this NOAA database significantly expanded in the mid-1990s and the analysis focuses on 15 of the more recent years (1996-2010) with more complete data.  A January 2006 California flood event mis-coded in the NOAA data as $115 billion impact was excluded.

Economic impact (as reflected by property/crop damage) is primarily caused by hurricanes, storm surges, and floods.  These three events collectively drive ~60% of property/crop damage.  

Human health impact (reflected by injuries/fatalities, with fatalities weighted more heavily) is heavily driven by extreme heat and tornadoes.  These two events account for ~40% of injuries and fatalities.

While there is occasional overlap among top causes, it appears that different weather events tend to create large economic impacts vs. large health impacts.  The NOAA dataset attributes relatively few deaths and injuries to hurricanes (and their derivatives such as storm surges), which may underly this finding.

## Data Processing  

### Initial Data Acquisition and Loading
Data for this project was obtained from [Storm Data Download](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2).  The raw file is downloaded to the working directory, write-protected, and loaded using R.  

```{r}
myCSV <- read.csv("repdata-data-StormData.csv.bz2",stringsAsFactors = FALSE,na.strings=c("NA",""))
str(myCSV)
```

Inspection of the file shows there are 902,297 observations across 37 variables.

### Filtering to maintain only relevant columns
Some columns are not relevant for this analysis.  We are particularly interested in event date "BGN_DATE"; event type "EVTYPE"; health impacts "FATALITIES" and "INJURIES"; and economic impacts "PROPDMG" and "CROPDMG".  Note that economic impacts are reported such that magnitude is contained in separate columns "PROPDMGEXP" and "CROPDMGEXP".

We also keep the STATE and REFNUM columns in case there are useful for follow-on analysis.

```{r}
myVars <- c("REFNUM","BGN_DATE","STATE","EVTYPE","FATALITIES","INJURIES",
            "PROPDMG","CROPDMG","PROPDMGEXP","CROPDMGEXP")
myFiltered <- myCSV[,myVars]
```

### Calculating the event year/decade and checking the resulting data
We also create a year and decade variable for future analysis and inspect the file contents.

```{r}
myFiltered$year <- as.numeric(format(as.Date(myFiltered$BGN_DATE,format="%m/%d/%Y"),"%Y"))
myFiltered$decade <- 10 * (myFiltered$year %/% 10)
str(myFiltered)
summary(myFiltered[,c(5,6,7,8,11,12)])
table(myCSV$PROPDMGEXP,myCSV$CROPDMGEXP,useNA="always")
```

The summary statistics suggest a succesful conversion to numeric year/decade, with 50% of the records being from 1995-2007.  There are no NA records in the numeric fields that we plan to use.  We maintained all of the 902,297 observations along with the 10 key variables and 2 new derived variables (year and decade).

### Calculating property/crop damage per event  
#### _Identify and plan around anomalous magnitudes for crop and property damage_  
Inspection of the PROPDMGEXP and CROPDMGEXP table above, along with exploration of the associated records, suggests K/k is for thousands, M/m is for millions, and B is for billions (e.g., huge hurricane).

There are 321 records with PROPDMGEXP symbols that cannot be interpreted and 27 records with CROPDMGEXP symbols that cannot be interpreted.  Manual inspection of a sample of these records shows them to be unremarkable and they are all treated as having $0 of damage.  These anomalies occured primarily in 1995.

Further, there are 76 records where PROPDMGEXP is NA while PROPDMG > 0; and 3 records where CROPDMGEXP is NA while CROPDMG > 0.  These anomalies all occur in 1993-1995 and inspection of a sample of these records shows them to be unremarkable.  These anomalies will be treated as having $0 of damage.

#### _Convert property and crop damage to dollars based on k/K or m/M or B_  
We convert all damage to millions of dollars for ease of future reporting.  If the PROPDMGEXP or CROPDMGEXP variable is not interpretable, it will stick with the default zero multiplier.  

```{r}
myFiltered$multProp <- 0
myFiltered$multProp[which(myFiltered$PROPDMGEXP %in% c("k","K"))] <- .001
myFiltered$multProp[which(myFiltered$PROPDMGEXP %in% c("m","M"))] <- 1
myFiltered$multProp[which(myFiltered$PROPDMGEXP %in% c("b","B"))] <- 1000

myFiltered$multCrop <- 0
myFiltered$multCrop[which(myFiltered$CROPDMGEXP %in% c("k","K"))] <- .001
myFiltered$multCrop[which(myFiltered$CROPDMGEXP %in% c("m","M"))] <- 1
myFiltered$multCrop[which(myFiltered$CROPDMGEXP %in% c("b","B"))] <- 1000

myFiltered$dollarProp <- myFiltered$PROPDMG * myFiltered$multProp
myFiltered$dollarCrop <- myFiltered$CROPDMG * myFiltered$multCrop

dim(myFiltered)
colnames(myFiltered)
set.seed(216160957)
myFiltered[sample(nrow(myFiltered),15,replace=FALSE),][c(1,8,10,13,15),]  ## Rows of some interest
myFiltered[myFiltered$multProp==1000,][sample(40,6,replace=FALSE),]
```

As expected, myFiltered still contains 902,297 observations.  The original 10 key variables plus year/decade remain.  Four new variables for calculating/reporting the dollar value of damage have been added.  

The 2011 flood in Vermont correctly converts 500M to 500, while records with k/K correctly divide by 1000.  A sampling of records with billions of reported property damage reveals catastrophic events commonly associated with reports of widespread property destruction.  Conversion to millions of dollars appears as intended in these samples.

Lastly, we examine the database totals for the key numeric variables (recall that the conversions for dollarCrop and dollarProp made each of these represent millions of dollars).  

```{r}
colSums(myFiltered[,c(5:8,15:16)])
```

Of note, the estimated $49 billion of crop damage and $427 billion of property damage over ~60 years is far more plausible than the unconverted $11 million of property damage and $1.4 million of crop damage.

### Aggregating event types for further analysis
#### _Declare an event type by finding key words in EVTYPE_

The EVTYPE variable is a bit messy, and we clean it using key word searches.  Exploratory analysis was run to understand key words associated with EVTYPES causing the strong majority of economic and health impacts.  The exploratory analysis is not shown for purposes of brevity.

A strategy to convert EVTYPE to eventType was developed, with the default being to declare eventType as "All Other" unless a specific keyword can be found.  The priority of the assignments is the reverse order of the variables declared in myShortEvent.  That is to say that if EVTYPE were "Thunderstorm with high wind due to tornado", then it would map to TORNADO and not to WIND or THUNDERSTORM.  The general intent was to prioritize more descriptive/severe event names over more general event names.

```{r}
myShortEvent <- c("WINTER","WIND","FREEZE","COLD","HEAT","RAIN","SNOW","ICE","HAIL","FLD","FLOOD",
                  "TSUNAMI","SURGE","DROUGHT","FOG","DUST","TSTM","THUNDERSTORM","LIGHTNING","FIRE",
                  "BLIZZARD","AVALANCHE","FLASH","SURF","RIP","TYPHOON","HURRICANE","TROPICAL","TORNADO"
                  )

myFiltered$eventType <- "All Other"

for (myMatch in myShortEvent) {
    myFiltered$eventType[grep(myMatch,myFiltered$EVTYPE,ignore.case=TRUE)] <- myMatch
}
```

Further, a few event types were consolidated as they appear highly related to one another.

```{r}
myFiltered$eventType[myFiltered$eventType %in% c("HURRICANE","TYPHOON")] <- "TROPICAL"
myFiltered$eventType[myFiltered$eventType=="SURF"] <- "RIP"
myFiltered$eventType[myFiltered$eventType=="TSTM"] <- "THUNDERSTORM"
myFiltered$eventType[myFiltered$eventType %in% c("FLD")] <- "FLOOD"
myFiltered$eventType[myFiltered$eventType %in% c("TSUNAMI")] <- "FLASH"
myFiltered$eventType[myFiltered$eventType=="FREEZE"] <- "COLD"
myFiltered$eventType[myFiltered$EVTYPE %in% c("NON TSTM WIND","NON-TSTM WIND")] <- "WIND"

```

#### _Declare an event class by further grouping eventType_
The event types were further aggregated to simplify some of the downstream reporting.

```{r}
myFiltered$masterType <- "Other"

myFiltered[myFiltered$eventType %in% c("HAIL","ICE","LIGHTNING","RAIN","SNOW",
                                       "THUNDERSTORM","TORNADO","TROPICAL","WINTER"),]$masterType <- "Storm"

myFiltered[myFiltered$eventType %in% c("COLD","DROUGHT","HEAT"),]$masterType <- "Temp/Dew Point"
myFiltered[myFiltered$eventType %in% c("BLIZZARD","WIND","DUST"),]$masterType <- "Wind"
myFiltered[myFiltered$eventType %in% c("FLASH","FLOOD","RIP","SURGE"),]$masterType <- "Water"
myFiltered[myFiltered$eventType %in% c("AVALANCHE","FIRE","FOG"),]$masterType <- "Assorted"
```

### Calculating a single number for economic impact, health impact, and total impact   
The focus of this analysis is to understand the impact of weather events on the economy and human health.  Additional variables are created to assess these overall impacts:  

* econImpact is the straight sum of crop damage and property damage, in millions of dollars
* healthImpact is an indexed aggregate of fatalities and injuries; the value of a human life lost is indexed as $5 million while the value of a human injury is indexed as $100,000; the sum of these indices is defined as healthImpact, in millions of dollars
* totalImpact is the straight sum of econImpact and healthImpact, in millions of dollars

```{r}
myFiltered$econImpact <- myFiltered$dollarProp + myFiltered$dollarCrop ## in millions
myFiltered$healthImpact <- 0.1 * myFiltered$INJURIES + 5 * myFiltered$FATALITIES ## in millions
myFiltered$totalImpact <- myFiltered$econImpact + myFiltered$healthImpact ## in millions
dim(myFiltered)
colSums(myFiltered[,c(5:6,15:16,19:21)])
print(paste0("Property damage represents ",round(100*sum(myFiltered[,15])/sum(myFiltered[,19]),0),
             "% of the $",round(sum(myFiltered[,19])/1000,0)," billion of indexed economic impact"
             )
      )
print(paste0("Fatalities represent ",round(5*100*sum(myFiltered[,5])/sum(myFiltered[,20]),0),
             "% of the $",round(sum(myFiltered[,20])/1000,0)," billion of indexed health impact"
             )
      )
print(paste0("Economic impact represents ",round(100*sum(myFiltered[,19])/sum(myFiltered[,21]),0),
             "% of the $",round(sum(myFiltered[,21])/1000,0)," billion of indexed total impact"
             )
      )
```

myFiltered continues to have all 902,297 observations.  In addition to the 16 variables previously carried, we have added descriptors for each event (eventType, masterType) and indexed millions of dollars of impact for each category (econImpact, healthImpact, totalImpact).  There are now 21 variables.

While there are many more injuries than fatalities, the assumption that loss of human life is of much greater impact than injury drives fatalities to represent 84% of the indexed health impact.  Property damage is of order of magnitude greater than crop damage, representing 90% of the indexed economic impact.  Economic impact is generally much greater than human impact, as expected of weather impacting highly-developed areas.  These relative impacts are sensitive to indexing assumptions for fatality at $5 million and injury at $100,000.  


## Results

### Can we use all years of NOAA data for our analysis?
#### _What is the pattern by year and impact type?_
In preparation for graphing, we aggregate the data by year and masterType.  Further, we brew a qualitative color palette to help with interpretation.  The RColorBrewer library is required.

```{r}
yearTotals <- aggregate(cbind(econImpact,healthImpact,totalImpact) ~ year + masterType,
                        data=myFiltered,FUN=sum
)

yearGraph <- reshape(yearTotals[,c(1,2,5)],v.names="totalImpact",timevar="masterType",
                     idvar="year",direction="wide")

yearGraph <- yearGraph[order(yearGraph$year),]

library(RColorBrewer)
myCol <- brewer.pal(6,"Accent")
```

A stacked bar chart reveals the data underwent a significant transition in 1993 when additional event types began to be tracked.

```{r}
par(las=2)
par(mar=c(5,5,4,2))
barplot(t(yearGraph[,c(4,6,5,7,2,3)])/1000,names.arg=yearGraph$year,col=myCol[6:1],cex.names=0.5)
legend("topleft",legend=c("Other","Misc","Wind","T/D","Water","Storm"),fill=myCol)
title(main="Index of Total Weather Impact by Year",ylab="Indexed Impact ($ billions)",xlab="Year")
```

There is significant spikiness to the total impact data, with pronounced peaks in:

* 2005 - mix of storm and water, revealed by further exploration to be driven by massive destruction attributed to Hurricane Katrina's August 2005 landfall on the LA and MS coastlines (with associated storm surge)
* 2006 - revealed by further exploration to be almost entirely driven by a single flooding event in CA in January 2006

The Katrina impact in this data is in the ballpark of the ~$100 billion of damage commonly associated to the storm.  Many of the ~2,000 deaths commonly attributed to Katrina are not in the database.  NOAA documentation notes death tracking can be finnicky, particularly when caused directly vs. indirectly by weather events.

The 2006 spike is obvious data error introduced by REFNUM 605943.  There was a flood in California, but with impact of ~$100 million.  Due to data entry error, this was coded as a $115 billion event.  Excluding  this record, the balance of 2006 has $13.8 billion of total impact, in line with other years.

#### _What do we do about it?_  
The remainder of this analysis will focus only on the years 1996-2010.  Data before 1993 is incomplete as per the above bar chart.  This document previously noted some anomalous characters for PROPDMGEXP and CROPDMGEXP in 1993-1995.  These three years not being vital to answering key questions, we exclude the years as possible teething issues associated with the methodology change.  Further, this database is of vintage November 2011, with attendant uncertainty as to the completeness and validation of records reported for 2011.

Further, REFNUM 605943 is deleted as it is clear data entry error with very large analysis impact.

```{r}
yearAnalyze <- subset(myFiltered,c(year >= 1996 & year <= 2010 & REFNUM != 605943) )
dim(yearAnalyze)
```

There are 591,335 observations remaining for analysis, including the 21 variables described previously.

Each REFNUM for the top-24 events by total-impact was examined in the raw myCSV file.  There is no other obvious outlier error and most of the data match to large tropical storms and/or major floods, both known to cause extensive property damage.  I am somewhat skeptical about reported impact in the billions of dollars for AZ hail event 834634, NM fire event 398999, CA fire event 488004, and TN flooding event 808257.  I lack domain expertise to delete these and therefore retained them for this analysis.

Please use caution in interpreting the remainder of this report.  The NOAA data may need further scrubbing to refine these initial findings.

```{r}
## REFNUM associated with the top-24 impact-causing events
yearAnalyze[order(-yearAnalyze$totalImpact),]$REFNUM[1:24]
```

### What events have the greatest impact?
#### _Final pre-processing and aggregation_  
The data were aggregated by eventType.  A mapping file was also applied to give more descritpive names to the weather event types.

```{r}
causeSum <- aggregate(cbind(FATALITIES,INJURIES,dollarProp,dollarCrop,econImpact,healthImpact,totalImpact) ~
                      eventType,data=yearAnalyze,FUN=sum)

myMap <- data.frame(origName = c("All Other","AVALANCHE","BLIZZARD","COLD",
                                 "DROUGHT","DUST","FIRE","FLASH",
                                 "FLOOD","FOG","HAIL","HEAT",
                                 "ICE","LIGHTNING","RAIN","RIP",
                                 "SNOW","THUNDERSTORM","TORNADO","TROPICAL",
                                 "WIND","WINTER","SURGE"),
                    modName = c("All Other","Avalanche","Blizzard","Extreme Cold",
                                "Drought","Dust Storm","Wildfire","Flash Flooding",
                                "Flood (non-flash)","Fog","Hail","Excessive Heat",
                                "Ice Storm","Lightning Strike","Rain","Rip Current/Surf",
                                "Snow","Thunderstorm","Tornado","Hurricane or TS",
                                "Wind","Wintry Mess","Storm Surge"),
                    stringsAsFactors=FALSE
)

for (strNames in myMap$origName){
    causeSum$descName[causeSum$eventType == strNames] <- myMap$modName[myMap$origName==strNames]
}

```

#### _Greatest economic impacts_  
The top-10 events by total economic impact are calculated, with a bar-plot produced to show the proportion of total 1996-2010 economic (crop and property damge) impact associated to that event.

Plainly, hurricanes, storm surges, and flooding cause the strong majority of economic impact.  While Katrina plays a big role in driving this, it is not uncommon for the US to experience major hurricance landfalls and extensive floods.  An insurance company or public safety agency would rightly see all as significant hazards.

```{r}

myPlot <- causeSum[order(-causeSum$econImpact),]

par(las=1)
par(mar=c(6,8,4,2))
barplot(myPlot$econImpact[10:1]/sum(myPlot$econImpact),
        names.arg=myPlot$descName[10:1],horiz=TRUE,
        col="blue",xlim=c(0,0.4)
)

title(main="Impact of top-10 (by economics) weather events in 1996-2010",
      xlab="% of economic impact caused (sum of crop damage and property damage)"
)

colSums(myPlot[1:10,2:8])/colSums(myPlot[,2:8])
```

These top-10 economic events account for ~95% of the economic impact but less than 50% of the health impact.

% Caused | Fatality | Injury | Property | Crop   | Economic | Health
---------|----------|--------|----------|--------|----------|--------
top-10   |  41%     |  66%   |   96%    |  88%   |   95%    |  44%


#### _Greatest health impacts_  
The top-10 events by total health impact are calculated, with a bar-plot produced to show the proportion of total 1996-2010 health (indexed fatality plus injury) impact associated to that event.

Excessive heat, tornadoes, and flash floods are the largest drivers of total health impact due to weather in 1996-2010.  Notably, of the top-3 economic drivers (hurricane, storm surge, flood), only floods make the top-10 for health impact.  This is likely related to the comment above about few deaths in this NOAA dataset being attributed to Katrina and similar hurricanes.

As a reminder, "flash flooding" is broken out separately from "flooding" (widespread river floods) and "storm surge".  Despite this, "flash flooding" ranks third in causing human health impact.

```{r}
myPlot <- causeSum[order(-causeSum$healthImpact),]

par(las=1)
par(mar=c(6,8,4,2))
barplot(myPlot$healthImpact[10:1]/sum(myPlot$healthImpact),
        names.arg=myPlot$descName[10:1],horiz=TRUE,
        col="orange",xlim=c(0,0.4)
)

title(main="Impact of top-10 weather events (by human health) in 1996-2010",
      xlab="% of injury-equivalents caused (each fatality treated as 50 injury-equivalents)"
)

colSums(myPlot[1:10,2:8])/colSums(myPlot[,2:8])
```

These top-10 health events account for ~85% of the health impact but less than 30% of the economic impact.

% Caused | Fatality | Injury | Property | Crop   | Economic | Health
---------|----------|--------|----------|--------|----------|---------
top-10   |  86%     |  85%   |   29%    |  32%   |   29%    |  86%

  
  
## Addendum #1 - EVTYPE mapping  
This addendum is to provide further description of the implication of aggregating raw EVTYPE in to more descriptive eventType. See below for the EVTYPE associated to each aggregate description.  Only EVTYPE totalling at least $99 million over 1996-2010 are shown.    

```{r}
testValue <- aggregate(cbind(totalImpact,econImpact,healthImpact,dollarProp,dollarCrop,FATALITIES,INJURIES) ~
                             eventType + EVTYPE,data=yearAnalyze,FUN=sum
                       )

for (strNames in myMap$origName){
    testValue$descName[testValue$eventType == strNames] <- myMap$modName[myMap$origName==strNames]
}

testValue <- testValue[order(testValue$descName,-testValue$totalImpact),]

testValue$totalImpact <- round(testValue$totalImpact,1)
testValue$dollarProp <- round(testValue$dollarProp,1)
testValue$dollarCrop <- round(testValue$dollarCrop,1)

testValue[testValue$totalImpact>=99,c(10,2:3,6:9)]
colSums(testValue[testValue$totalImpact>=99,c(3,6:9)])/colSums(testValue[testValue$totalImpact>=0,c(3,6:9)])
```

The EVTYPE shown above account for 97%-99% of the impact of the various events.


## Addendum #2 - Top-10 events for each type of individual impact  
The raw data are such that the indexing and aggregation methodologies lead crop damage and injuries to take a back-seat to property damage and fatalities.

Below are the top-10 EVTYPE for each of dollarCrop, INJURIES, dollarProp, and FATALITIES.  Drought in particular drives a lot of crop damage, but crop damage is just 10% of economic damge and since drought causes negligible property damage, it never appears as a top cause in this report.

```{r}
for (myString in c("dollarCrop","INJURIES","dollarProp","FATALITIES")) {
    myLoc <- grep(myString,colnames(testValue))
    testValue <- testValue[order(-testValue[,myLoc]),]
    print("--------------------")
    print(paste0("----- Data for top-10 drivers of: ",myString))
    print(testValue[1:10,c(2,10,3,6:9)])
    print("--------------------")
    print(paste0("These top 10 EVTYPE for: ",myString," represent the below proportions of total"))
    print(colSums(testValue[1:10,c(3,6:9)])/colSums(testValue[,c(3,6:9)]))
}
```

