---
title: "Economic and Health Impact of US Weather Events in 1996-2010"
author: "davegoblue"
date: "February 19, 2016"
output: html_document
---

## Synopsis  
This document summarizes the economic and health impact of US weather events as captured in an NOAA database spanning 1950-2011.  Weather event tracking in this NOAA database significantly expanded in the mid-1990s and the analysis focuses on 15 years (1996-2010) with more complete data.  The raw dataset is known to be only a sample of the full NOAA Storm Data (e.g., it does not capture Louisiana fatalaties from Hurricane Katrina), and findings should be interpreted with that in mind.

* Economic impact (as reflected by property/crop damage) is primarily caused by hurricane winds, storm surges, and floods.  These three events collectively cause ~60% of property/crop damage.  

* Human health impact (as reflected by injuries/fatalities, with fatalities weighted more heavily) is heavily driven by extreme heat and tornadoes.  These two events combine to cause ~40% of injuries and fatalities.  

While there is occasional overlap among top causes, it appears different types of weather events tend to create large economic impacts vs. large health impacts.  
  
## Data Processing  
  
###_Initial Data Acquisition and Loading_  
Data for this project was obtained from [Storm Data Download](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2).  The raw ZIP file is downloaded to the working directory, write-protected, and loaded using R.  The directions for this assignment asked that we start from the ZIP file after having downloaded it to our machine.

```{r, cache=TRUE}
## read.csv() directly reads .csv.bz2 files, managing decompression as part of the function
myCSV <- read.csv("repdata-data-StormData.csv.bz2", stringsAsFactors = FALSE, na.strings=c("NA",""))
str(myCSV)
```

Inspection of the file shows 902,297 observations and 37 variables.  

###_Filtering to maintain only relevant columns_  
Some columns are not relevant for this analysis.  We are particularly interested in event date "BGN_DATE"; event type "EVTYPE"; health impacts "FATALITIES" and "INJURIES"; and economic impacts "PROPDMG" and "CROPDMG".  Note that economic impacts are reported such that order of magnitude (thousands, millions, or billions) is contained in separate columns "PROPDMGEXP" and "CROPDMGEXP".

We also keep the STATE and REFNUM columns in case there are useful for QC or follow-on analysis.

```{r}
myVars <- c("REFNUM","BGN_DATE","STATE","EVTYPE","FATALITIES","INJURIES",
            "PROPDMG","CROPDMG","PROPDMGEXP","CROPDMGEXP")
myFiltered <- myCSV[,myVars]
```

###_Calculating the event year/decade and checking the resulting data_  
We also create a year and decade variable for future analysis and inspect the file contents.

```{r}
myFiltered$year <- as.numeric(format(as.Date(myFiltered$BGN_DATE,format="%m/%d/%Y"),"%Y"))
myFiltered$decade <- 10 * (myFiltered$year %/% 10)
str(myFiltered)
summary(myFiltered[,c(5,6,7,8,11,12)])
```

The summary statistics suggest a succesful conversion to numeric year/decade, with 50% of the records being from 1995-2007.  There are no NA records in the numeric fields that we plan to use.  We maintained all of the 902,297 observations along with the 10 key variables and 2 new derived variables (year and decade).

###_Calculating property/crop damage per event_  
##### **_Identify and plan around anomalous magnitudes for crop and property damage_**  
A contingency table is created to understand the entries in PROPDMGEXP and CROPDMGEXP:  
```{r}
table(myCSV$PROPDMGEXP,myCSV$CROPDMGEXP,useNA="always")
```
  
Inspection of the PROPDMGEXP and CROPDMGEXP table, along with exploration of the associated records, suggests K/k is for thousands, M/m is for millions, and B is for billions (e.g., very large hurricanes and floods).

There are 321 records with PROPDMGEXP symbols that cannot be interpreted and 27 records with CROPDMGEXP symbols that cannot be interpreted.  Manual inspection of a sample of these records shows them to be unremarkable and they are all treated as having $0 of damage.  These anomalies occured primarily in 1995.

Further, there are 76 records where PROPDMGEXP is NA while PROPDMG > 0; and 3 records where CROPDMGEXP is NA while CROPDMG > 0.  Maunal inspection of a sample of these records shows them to be unremarkable, and they are all treated as having $0 of damage.  These anomalies all occur in 1993-1995.  

##### **_Convert property and crop damage to dollars based on k/K or m/M or B_**  
We convert all damage to millions of dollars for ease of future reporting.  If the PROPDMGEXP or CROPDMGEXP variable is not interpretable, it will stick with the default zero multiplier (treat as $0).  

```{r}
myFiltered$multProp <- 0
myFiltered$multProp[which(myFiltered$PROPDMGEXP %in% c("k","K"))] <- .001  ## thousands to millions
myFiltered$multProp[which(myFiltered$PROPDMGEXP %in% c("m","M"))] <- 1 ## leave millions as is
myFiltered$multProp[which(myFiltered$PROPDMGEXP %in% c("b","B"))] <- 1000 ## billions to millions

myFiltered$multCrop <- 0
myFiltered$multCrop[which(myFiltered$CROPDMGEXP %in% c("k","K"))] <- .001  ## thousands to millions
myFiltered$multCrop[which(myFiltered$CROPDMGEXP %in% c("m","M"))] <- 1 ## leave millions as is
myFiltered$multCrop[which(myFiltered$CROPDMGEXP %in% c("b","B"))] <- 1000 ## billions to millions

myFiltered$dollarProp <- myFiltered$PROPDMG * myFiltered$multProp ## field now in millions
myFiltered$dollarCrop <- myFiltered$CROPDMG * myFiltered$multCrop ## field now in millions

dim(myFiltered)
colnames(myFiltered)
set.seed(216160957)
myFiltered[sample(nrow(myFiltered),15,replace=FALSE),][c(1,8,10,13,15),]  ## Rows of more interest
myFiltered[myFiltered$multProp==1000,][sample(40,6,replace=FALSE),]
```

As expected, myFiltered still contains 902,297 observations.  There are 16 variables (original 10 key variables plus year/decade plus four new variables for use in the dollar value of damage as just described).  

The 2011 flood in Vermont correctly converts 500M to 500, while records with k/K correctly divide by 1000.  A sampling of records with billions of reported property damage reveals catastrophic event types commonly associated with reports of widespread property destruction.  Conversion to millions of dollars appears as intended in these samples.

Lastly, we examine the database totals for the key numeric variables (recall that conversions for dollarCrop and dollarProp made each of these represent millions of dollars).  

```{r}
colSums(myFiltered[,c(5:8,15:16)])
```

Of note, the estimated $427 billion of property damage and $49 billion of crop damage over ~60 years is far more plausible than the unconverted $11 million of property damage and $1.4 million of crop damage.

###_Aggregating event types for further analysis_  
##### **_Declare an event type by finding key words in EVTYPE_**  

The EVTYPE variable is a bit messy, and we clean it with key word searches.  Exploratory analysis was run to understand key words associated with EVTYPES causing the strong majority of economic and health impacts.  The main implications of the approach arising from that exploratory analysis are included below as Addendum #1.  

A strategy to convert EVTYPE to eventType was developed, with the default being to declare eventType as "All Other" unless a specific keyword can be found.  The priority of the assignments is the reverse order of the variables declared in myShortEvent.  That is to say that if EVTYPE were "Thunderstorm with high wind due to tornado", then it would map to TORNADO and not to WIND or THUNDERSTORM.  The general intent was to prioritize more descriptive/severe event names over more general event names.

```{r}
myShortEvent <- c("WINTER","WIND","FREEZE","COLD","HEAT","RAIN","SNOW","ICE","HAIL","FLD","FLOOD",
                  "TSUNAMI","SURGE","DROUGHT","FOG","DUST","TSTM","THUNDERSTORM","LIGHTNING","FIRE",
                  "BLIZZARD","AVALANCHE","FLASH","SURF","RIP","TYPHOON","HURRICANE","TROPICAL","TORNADO"
                  )

myFiltered$eventType <- "All Other"

## This loop keeps overwriting eventType, so the last item from myShortEvent
## found within EVTYPE for a given record prevails; not a case sensitive search
for (myMatch in myShortEvent) {
    myFiltered$eventType[grep(myMatch,myFiltered$EVTYPE,ignore.case=TRUE)] <- myMatch
}
```

Further, a few event types were consolidated as they appear highly related to one another.

```{r}
myFiltered$eventType[myFiltered$eventType %in% c("HURRICANE","TYPHOON")] <- "TROPICAL"
myFiltered$eventType[myFiltered$eventType=="SURF"] <- "RIP"
myFiltered$eventType[myFiltered$eventType=="TSTM"] <- "THUNDERSTORM"
myFiltered$eventType[myFiltered$eventType %in% c("FLD")] <- "FLOOD"
myFiltered$eventType[myFiltered$eventType %in% c("TSUNAMI")] <- "FLASH"
myFiltered$eventType[myFiltered$eventType=="FREEZE"] <- "COLD"
myFiltered$eventType[myFiltered$EVTYPE %in% c("NON TSTM WIND","NON-TSTM WIND")] <- "WIND"

```

##### **_Declare an event class by further grouping eventType_**  
The event types were further aggregated to simplify some downstream reporting.

```{r}
myFiltered$masterType <- "Other"

myFiltered[myFiltered$eventType %in% c("HAIL","ICE","LIGHTNING","RAIN","SNOW",
                                       "THUNDERSTORM","TORNADO","TROPICAL","WINTER"),]$masterType <- "Storm"

myFiltered[myFiltered$eventType %in% c("COLD","DROUGHT","HEAT"),]$masterType <- "Temp/Dew Point"
myFiltered[myFiltered$eventType %in% c("BLIZZARD","WIND","DUST"),]$masterType <- "Wind"
myFiltered[myFiltered$eventType %in% c("FLASH","FLOOD","RIP","SURGE"),]$masterType <- "Water"
myFiltered[myFiltered$eventType %in% c("AVALANCHE","FIRE","FOG"),]$masterType <- "Assorted"
```

### Calculating a single number for economic impact, health impact, and total impact   
The focus of this analysis is to understand the impact of weather events on the economy and human health.  Additional variables are created to assess these overall impacts:  

* econImpact is the straight sum of crop damage and property damage, in millions of dollars
* healthImpact is an indexed aggregate of fatalities and injuries; the value of a human life lost is indexed as $5 million while the value of a human injury is indexed as $100,000; the sum of these indices is defined as healthImpact, in millions of dollars
* totalImpact is the straight sum of econImpact and healthImpact, in millions of dollars

```{r}
myFiltered$econImpact <- myFiltered$dollarProp + myFiltered$dollarCrop ## in millions
myFiltered$healthImpact <- 0.1 * myFiltered$INJURIES + 5 * myFiltered$FATALITIES ## indexed to millions
myFiltered$totalImpact <- myFiltered$econImpact + myFiltered$healthImpact ## indexed in millions

dim(myFiltered)
colSums(myFiltered[,c(5:6,15:16,19:21)])

print(paste0("Property damage represents ",round(100*sum(myFiltered[,15])/sum(myFiltered[,19]),0),
             "% of the $",round(sum(myFiltered[,19])/1000,0)," billion of indexed economic impact"
             )
      )
print(paste0("Fatalities represent ",round(100*5*sum(myFiltered[,5])/sum(myFiltered[,20]),0),
             "% of the $",round(sum(myFiltered[,20])/1000,0)," billion of indexed health impact"
             )
      )
print(paste0("Economic impact represents ",round(100*sum(myFiltered[,19])/sum(myFiltered[,21]),0),
             "% of the $",round(sum(myFiltered[,21])/1000,0)," billion of indexed total impact"
             )
      )
```

myFiltered continues to have all 902,297 observations.  In addition to the 16 variables previously carried, we have added descriptors for each event (eventType, masterType) along with indexed millions of dollars of impact (econImpact, healthImpact, totalImpact).  There are now 21 variables.  

While there are many more injuries than fatalities, the assumption that loss of human life is of much greater impact than injury drives fatalities to represent 84% of indexed health impact.  Property damage is of order of magnitude greater than crop damage, representing 90% of indexed economic impact.  Economic impact is generally much greater than human impact, with the relative impacts being sensitive to indexing assumptions for fatality at $5 million and injury at $100,000.  

Please see Addendum #2 below if interested in the event types having the greatest impact on any metric (cropDamage, propDamage, INJURIES, FATALITIES) on a standalone basis.  
  
  
## Results  
### _Can we use all years of NOAA data for our analysis?_  
##### **_What is the pattern by year and impact type?_**  
We aggregate the data by year and masterType and brew a qualitative color palette to help with interpretation (note that the RColorBrewer library is required).  A stacked bar chart for indexed impact per year by class of event is created:  

```{r}

## Approach for creating stacked bar charts adapted from
## http://www.r-bloggers.com/stacked-bar-charts-in-r/

yearTotals <- aggregate(cbind(econImpact,healthImpact,totalImpact) ~ year + masterType,
                        data=myFiltered,FUN=sum
)

yearGraph <- reshape(yearTotals[,c(1,2,5)],v.names="totalImpact",timevar="masterType",
                     idvar="year",direction="wide")

yearGraph <- yearGraph[order(yearGraph$year),]
str(yearGraph)

library(RColorBrewer)
myCol <- brewer.pal(6,"Accent")

par(las=2)
par(mar=c(4,4,2,1))

barplot(t(yearGraph[,c(4,6,5,7,2,3)])/1000,names.arg=yearGraph$year,col=myCol[6:1],cex.names=0.7)
legend("topleft",legend=c("Other","Misc","Wind","Temp/Humid","Water","Storm"),fill=myCol)
title(main="Index of Total Impact by Weather Type by Year",
      ylab="Indexed Impact ($ billions)",xlab="Year"
      )
```
  
**_Figure 1: US Weather Event Type Tracking Increased in 1993_**  
  
The stacked bars reveal a significant transition in 1993 when additional event types began to be tracked.  There is also spikiness to the total impact by year, with pronounced peaks in:  

* 2005 - mix of storm and water, revealed by further exploration to be driven by massive destruction attributed to Hurricane Katrina's August 2005 landfall on the LA and MS coastlines (with associated storm surge)
* 2006 - revealed by further exploration to be almost entirely driven by a single flooding event in CA in January 2006

The Katrina impact in this data is in the ballpark of the ~$100 billion of damage commonly associated to the storm.  Many of the ~2,000 deaths commonly attributed to Katrina are not in the database, particularly those in the state of Louisiana (LA).  In response to a query, we learned that data provided for this analysis is deliberately a subset of the full NOAA Storm Data, and that lack of Katrina deaths in LA is expected. 

The 2006 spike is obvious data error introduced by REFNUM 605943.  There was a flood in California, but with impact of ~$100 million.  Due to data entry error, this was coded as a $115 billion event.  Other than  this record, the balance of 2006 has ~$13 billion of total impact, in line with nearby years.

##### **_What do we do about it?_**  
The remainder of this analysis will focus only on the years 1996-2010.  Data before 1993 is incomplete as per the above chart.  This document previously noted anomalies for PROPDMGEXP and CROPDMGEXP in 1993-1995.  These three years not being vital to answering key questions, we exclude the years as possible teething issues associated with a methodology change.  Further, this database is of vintage November 2011, with attendant uncertainty as to the completeness and validation of records reported for 2011.

In addition, REFNUM 605943 is deleted as it is clear data entry error with very large analysis impact.

```{r}
yearAnalyze <- subset(myFiltered,c(year >= 1996 & year <= 2010 & REFNUM != 605943) )
dim(yearAnalyze)
print(aggregate(cbind(totalImpact=round(totalImpact,0),
                      econImpact=round(econImpact,0),
                      healthImpact=round(healthImpact,0)
                      ) 
                ~ year, data=yearAnalyze, FUN=sum
                )
      )
```

There are 591,335 observations remaining for analysis, including the 21 variables described previously.

Each REFNUM for the top-24 events by total-impact was examined in the raw myCSV file.  There is no other obvious outlier error and many of the data match to large tropical storms and/or major floods (both known to cause extensive property damage).  I am somewhat skeptical about reported impact in the billions of dollars for AZ hail event 834634, NM fire event 398999, CA fire event 488004, and TN flooding event 808257.  I lack domain expertise to delete these and therefore retained them for this analysis.

Please use caution in interpreting the remainder of this report.  The dataset provided is known to be merely a sample of the NOAA Storm Data and may also need further scrubbing for some of the largest impacts.  
  
```{r}
## For reference, these are the REFNUM for the top-24 impact-causing events
yearAnalyze[order(-yearAnalyze$totalImpact),]$REFNUM[1:24]
```

### _What events have the greatest impact?_  
##### **_Final pre-processing and aggregation_**  
The data were aggregated by eventType.  A mapping file was also applied to give more descritpive names to the weather event types.

```{r}
causeSum <- aggregate(cbind(FATALITIES,INJURIES,dollarProp,dollarCrop,econImpact,healthImpact,totalImpact) ~
                      eventType,data=yearAnalyze,FUN=sum)

myMap <- data.frame(origName = c("All Other","AVALANCHE","BLIZZARD","COLD",
                                 "DROUGHT","DUST","FIRE","FLASH",
                                 "FLOOD","FOG","HAIL","HEAT",
                                 "ICE","LIGHTNING","RAIN","RIP",
                                 "SNOW","THUNDERSTORM","TORNADO","TROPICAL",
                                 "WIND","WINTER","SURGE"),
                    modName = c("All Other","Avalanche","Blizzard","Extreme Cold",
                                "Drought","Dust Storm","Wildfire","Flash Flooding",
                                "Flood (non-flash)","Fog","Hail","Excessive Heat",
                                "Ice Storm","Lightning Strike","Rain","Rip Current/Surf",
                                "Snow","Thunderstorm Wind","Tornado","Hurricane Wind",
                                "Wind (non-TSTM/Hurricane)","Wintry Mess","Storm Surge"),
                    stringsAsFactors=FALSE
)

for (strNames in myMap$origName){
    causeSum$descName[causeSum$eventType == strNames] <- myMap$modName[myMap$origName==strNames]
}

```

##### **_Greatest economic impacts_**  
The top-10 events by total economic impact are calculated, with a bar-plot produced to show the proportion of total 1996-2010 economic (crop and property damge) impact associated to that event.

Hurricane winds, storm surges (typically due to hurricanes), and flooding cause the strong majority of economic impact.  While Katrina plays a big role in driving this, it is not uncommon for the US to experience major hurricance landfalls and widespread river floods.  An insurance company or public safety agency would rightly see these as significant hazards to the local economy.

```{r}

myPlot <- causeSum[order(-causeSum$econImpact),]

par(las=1)
par(mar=c(6,12,4,2))

barplot(myPlot$econImpact[10:1]/sum(myPlot$econImpact),
        names.arg=myPlot$descName[10:1],horiz=TRUE,
        col="blue",xlim=c(0,0.4),xaxt="n"
)

axis(1,at=seq(0,0.4,by=0.1),lab=paste0(pretty(100*seq(0,0.4,by=0.1)),"%"))

title(main="Top-10 (by economic impact) event types: 1996-2010",
      xlab="% of economic impact (crop damage plus property damage)"
)

```
  
**_Figure 2: US Weather Event Types Causing Greatest Economic Impact (1996-2010)_**  
  
```{r}
colSums(myPlot[1:10,2:8])/colSums(myPlot[,2:8])
```

These top-10 economic events account for ~95% of the economic impact but less than 50% of the health impact.

% Caused | Fatality | Injury | Property | Crop   | Economic | Health
---------|----------|--------|----------|--------|----------|--------
top-10   |  41%     |  66%   |   96%    |  88%   |   95%    |  44%


##### **_Greatest health impacts_**  
The top-10 events by total health impact are calculated, with a bar-plot produced to show the proportion of total 1996-2010 health (indexed fatality plus injury) impact associated to that event.

Excessive heat, tornadoes, and flash floods are the largest drivers of total health impact due to weather in 1996-2010.  Notably, of the top-3 economic drivers (hurricane winds, storm surges, floods), only floods make the top-10 for health impact.  This is likely at least in part related to the comment above about the dataset being a sample (e.g., few deaths in this dataset being attributed to Katrina).  

As a reminder, "flash flooding" is broken out separately from "flooding" (typically widespread river floods) and "storm surge" (typically due to hurricanes).  Despite this, "flash flooding" ranks third in causing human health impact.  

```{r}
myPlot <- causeSum[order(-causeSum$healthImpact),]

par(las=1)
par(mar=c(6,12,4,2))

barplot(myPlot$healthImpact[10:1]/sum(myPlot$healthImpact),
        names.arg=myPlot$descName[10:1],horiz=TRUE,
        col="orange",xlim=c(0,0.4), xaxt="n"
)

axis(1,at=seq(0,0.4,by=0.1),lab=paste0(pretty(100*seq(0,0.4,by=0.1)),"%"))

title(main="Top-10 (by health impact) event types: 1996-2010",
      xlab="% of injury-equivalents (each fatality treated as 50 injury-equivalents)"
)

```
  
**_Figure 3: US Weather Event Types Causing Greatest Health Impact (1996-2010)_**  
  
```{r}

colSums(myPlot[1:10,2:8])/colSums(myPlot[,2:8])
```
  
These top-10 health events account for ~85% of the health impact but less than 30% of the economic impact.

% Caused | Fatality | Injury | Property | Crop   | Economic | Health
---------|----------|--------|----------|--------|----------|---------
top-10   |  86%     |  85%   |   29%    |  32%   |   29%    |  86%

  
Different types of events appear to cause different types of impacts (economic vs. health) in this sample of the NOAA Storm Data from 1996-2010.  
  
  
## Addendum
#### _Addendum #1 - EVTYPE mapping_  
This addendum is to provide further description of the implication of aggregating raw EVTYPE in to more descriptive eventType. See below for the EVTYPE associated to each aggregate description.  Only EVTYPE totalling at least $99 million over 1996-2010 are shown.    

```{r}
testValue <- aggregate(cbind(totalImpact,econImpact,healthImpact,dollarProp,dollarCrop,FATALITIES,INJURIES) ~
                             eventType + EVTYPE,data=yearAnalyze,FUN=sum
                       )

for (strNames in myMap$origName){
    testValue$descName[testValue$eventType == strNames] <- myMap$modName[myMap$origName==strNames]
}

testValue <- testValue[order(testValue$descName,-testValue$totalImpact),]

testValue$totalImpact <- round(testValue$totalImpact,1)
testValue$dollarProp <- round(testValue$dollarProp,1)
testValue$dollarCrop <- round(testValue$dollarCrop,1)

testValue[testValue$totalImpact>=99,c(10,2:3)]
colSums(testValue[testValue$totalImpact>=99,c(3,6:9)])/colSums(testValue[testValue$totalImpact>=0,c(3,6:9)])
```

The EVTYPE shown above account for 97%-99% of the various impacts of interest.


#### _Addendum #2 - Top-10 events for each type of impact on a stand-alone basis_  
The raw data are such that the indexing and aggregation methodologies lead crop damage and injuries to take a back-seat to property damage and fatalities.

Below are the top-10 EVTYPE for each of dollarCrop, INJURIES, dollarProp, and FATALITIES.  Drought in particular drives a lot of crop damage, but not much else.

```{r}
for (myString in c("dollarCrop","INJURIES","dollarProp","FATALITIES")) {
    myLoc <- grep(myString,colnames(testValue))
    testValue <- testValue[order(-testValue[,myLoc]),]
    print("--------------------")
    print(paste0("----- Data for top-10 drivers of: ",myString))
    print(testValue[1:10,c(2,10,3,6:9)])
    print("--------------------")
    print(paste0("These top 10 EVTYPE for: ",myString," represent the below proportions of total"))
    print(colSums(testValue[1:10,c(3,6:9)])/colSums(testValue[,c(3,6:9)]))
}
```

